Refinements for open automata

### 2022-07-13 2021-Refinement/paper/refinementarticle.tex
meeting notes

@@

This section presents our notations and the principles of automata. Except for minor changes in the notations, compared to previous works~\cite{pnets} the only new contribution is the definition of a composition operator for open automata.
%
%Notations will be defined with the operator \(\defnotation\) and names are given with the operator \(\defobject\) as follows:
%\begin{align*}
%	\mathit{notation\_with\_variables} & \defnotation \mathit{notated\_object\_using\_the\_variables} \\
%	\mathit{name} & \defobject \mathit{fully\_defined\_mathematical\_object}
%\end{align*}

@@

@@

\paragraph{Open automaton composition}

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the state of the automaton, i.e. if the variables of the OA verify a certain formula. 
In other words, for each pair of states, a boolean formula that may refer to the  variables of each of the two OA states whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

\paragraph{A bisimulation for open automata}

Bisimulation between OA was defined in~\cite{fh-bisim}. Instead of recalling its definition we show below the principles of this bisimulation.
\TODO{Je ne comprends pas cette figure... elle est incomplete ? Ou alors il faut un peu d'explications...}

@@

%The idea is that 
We rely on a classical notion of simulation inspired from~\cite{fh-bisim}. The idea is that two states related by a given relation and states that if one state can do a transition, then the other can do a transition too. But to limit the creation of deadlocks when simulating an automaton, we first define when a relation is preserved along identical transitions before characterising a first simple refinement relation  to compare automata with identical holes.

@@

\medskip

We define here simulation between labelled transition systems to Hole-equal simulation  between  open automata that have exactly the same holes.
\begin{definition}[Hole-equal simulation]
Consider two OAs \(\OAg[1]\) and \(\OAg[2]\) with  \(J_1 = J_2\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $S_1$ to $S_2$ if the following conditions hold: 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 = \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} = \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[10pt]
		\quad \mpar{\forall x \in X, J'_{2x}= J'_1} \nwedge\\[2pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\
\hspace{5em} \left( R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x}} \beta_{1j} = \beta_{2xj} \nwedge\\[12pt]
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}\right)
	\end{array} 
%\\
%	\wedge \mpar{V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\item[(3)] $R$ is  is preserved along identical transitions of \(A_1\) and \(A_2\).
\end{definition}

Note in this definition, instead of matching an instantiated transition (with all variables assigned) of the first automata to another instantiated transition of the second, it matches an open transition $t_1$ to a family of covering open transitions $t_{2x}^{x\in X}$. %Note also the second part of the conjunction of condition (2) expresses the deadlock reduction between automata (as introduced in Definition \ref{def:dpwd}).   
\TODO{The first and second conditions coincide with the natural way to prove inductively that an automaton simulates another by starting with the initial state. The third condition ensures that the refinement relation prevents the introduction of deadlocks.}

@@

\begin{figure}[h]
 \centering
   \includegraphics[width=.9\textwidth]{Figures/databaseRefine.pdf}
   \caption{The refined Open Automaton of the automaton of Fig.\ref{Fig:SpecOA}.  Two other actions,  in addition to  \texttt{qry} and \texttt{upd},  \texttt{back} representing a request to make a backup copy and    \texttt{cpy} is used to reflect the actual operation of copying. \label{Fig:RefineOA}}
\end{figure}

\TODO{The refinement relation  also specifies that the refined process must follow the behaviour of the specification, not only at every compatible state but also along any trace.}

 We will show in Section~\ref{sec:prop} that this refinement relation has good properties in terms of transitivity, compositionality, reflexivity, etc.

@@

\TODO{add an example}

@@

This definition leads to a notion of simulation as refinement the same as the one used in LOTOS \cite{Brinksma:87} and presented under the name extension refinement.
The extension consists in adding new functionalities by preserving existing ones. Indeed, in case there is no hole and no guard, i.e., in case the definitions of open automata coincide with those of automata, the relation allows new behaviour to be added in terms of new traces, but doesn't allow new deadlocks to
be added, i.e., for traces of the specification (abstract automaton) the implementation (refined automaton) must not add
deadlocks.

@@

\begin{lemma}[Tracked holes] By construction, if an automaton is the refinement of another one, it is also a refinement by tracking less holes.
\[\wrel{A_1}{A_2}{H} \land H'\subseteq H \implies \wrel{A_1}{A_2}{H'}\]
\end{lemma}
\TODO{example}

@@

\begin{center}
\includegraphics[scale=0.7]{Figures/Theorem2bis}
\end{center}

@@

% TODO: why is there several ovoid stacked only on one side?
\begin{center}
\includegraphics[scale=0.7]{Figures/Theorem1}
\end{center}

@@

\TODO{1 page}

@@

Additionally, because  the domains of the substitution function are disjoint, the substitution function has an effect only on the related elements,  we get:
\begin{align*}
& \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} 
%\wedge \mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}
\\
\implies & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}}  \land \mpar{\,\,\bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }\\
\implies & \mpar{R\mpar{s_{01}, s_{02}} \wedge  \bigwedge_\subbox{v_3\in V_3}  v^1_3=v^2_3\wedge s^1_3 = s^2_3} \psubst{ \sigma_{01} \uplus \sigma_{02} \uplus \sigma^1_{03}\uplus \sigma^2_{03}\uplus \sigma_{013} }\\
\implies & \sigma_{013} \uplus \sigma_{023} \vdash R'\mpar{s_{013}, s_{023}}
\end{align*}
\item Second, we need to prove for any OT $t_{13}$ in $T_{13}$ originating from $s_{13}$:
\begin{align*}
		&  \OTx{13}{}{13}{13} \in \fOT{s_{13}}
\end{align*}		
there exists an indexed family $t_{23x}$ of OTs originating from $s_{23}$ that simulate it: 
\begin{align*}
		&  \mpar{\OTx{23}{x}{23x}{23x} \in \fOT{s_{23}}}^{x \in X}
\end{align*}	
\begin{align*}		
		& \text{such that } \mpar{\forall x \in X, J'_{23x} \cap H' = J'_{13} \cap H'} \text{ and }\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash R'\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H'} \beta_{13j} = \beta_{23xj} \wedge\\[12pt]
			 g_{23x} \wedge R'\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}
Recall that by definition of composition and open automata refinement we have: 
\begin{align*}
&V_{13}= V_1\uplus V^1_3 \text{ and }
V_{23}= V_2\uplus V^2_3\\
& H' \subseteq J_3\uplus \mpar{J_1 \cap J_2} \setminus \{k\} = \mpar{J_3\uplus J_1 \setminus \{k\} } \cap \mpar{J_3\uplus J_2 \setminus \{k\}} \\
\end{align*}
First of all, we have by hypothesis $\wrel{A_1}{A_2}{H}$, then for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{2}$: 
\begin{align*}
\mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_{2}}}^{x \in X} 
\end{align*}
such that $\forall x \in X, J'_{2x} \cap H = J'_1 \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\ \mpar{R\mpar{s_1, s_{2}} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}}  \qquad \quad (*)
\end{multline*}	
\smallskip

@@

We reason on the way the OT $t_{23}$ has been obtained: either the hole $k$ of $A_2$ is involved in the transition or not.

@@

\item Second, we need to prove for any OT $t_{12}$ in $T_{12}$ originating from $s_{12}$

@@

We reason on the way the OT $t_{13}$ has been obtained: either the hole $k$ of $A_1$ is involved in the transition or not.

>>>>

\TODO{Entete : rabea}

\TODO{explication bisim: Ludo}

@@

This section presents our notations and the principles of automata. Except for minor changes in the notations, compared to previous works~\cite{fhbisim} the only new contribution is the definition of a composition operator for open automata.
%
%Notations will be defined with the operator \(\defnotation\) and names are given with the operator \(\defobject\) as follows:
%\begin{align*}
%	\mathit{notation\_with\_variables} & \defnotation \mathit{notated\_object\_using\_the\_variables} \\
%	\mathit{name} & \defobject \mathit{fully\_defined\_mathematical\_object}
%\end{align*}

@@

We use two operators to access pieces of information of the OA.

@@

\paragraph{Open automaton composition.}

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the state of the automaton, i.e. if the variables of the OA verify a certain formula. 
In other words, for each pair of states, a boolean formula that may refer to the  variables of each of the two OA states whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

\paragraph{A bisimulation for open automata.}

Bisimulation between OA was defined in~\cite{fhbisim}. Instead of recalling its definition we show below the principles of this bisimulation.
\TODO{Je ne comprends pas cette figure... elle est incomplete ? Ou alors il faut un peu d'explications...}

@@

%The idea is that 
We rely on a classical notion of simulation inspired from~\cite{fhbisim}. The idea is that two states related by a given relation and states that if one state can do a transition, then the other can do a transition too. But to limit the creation of deadlocks when simulating an automaton, we first define when a relation is preserved along identical transitions before characterising a first simple refinement relation  to compare automata with identical holes.

@@

\TODO{[LUDO]
We define here simulation between labelled transition systems to Hole-equal simulation  between  open automata that have exactly the same holes. }
\begin{definition}[Hole-equal simulation]
Consider two OAs \(\OAg[1]\) and \(\OAg[2]\) with  \(J_1 = J_2\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $S_1$ to $S_2$ if the following conditions hold: 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 = \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} = \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[10pt]
		\quad \mpar{\forall x \in X, J'_{2x}= J'_1} \nwedge\\[2pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\
\hspace{5em} \left( R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x}} \beta_{1j} = \beta_{2xj} \nwedge\\[12pt]
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}\right)
	\end{array} 
%\\
%	\wedge \mpar{V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\item[(3)] $R$ is  is preserved along identical transitions of \(A_1\) and \(A_2\).
\end{definition}

Note in this definition, instead of matching an instantiated transition (with all variables assigned) of the first automata to another instantiated transition of the second, it matches an open transition $t_1$ to a family of covering open transitions $t_{2x}^{x\in X}$. %Note also the second part of the conjunction of condition (2) expresses the deadlock reduction between automata (as introduced in Definition \ref{def:dpwd}).   
The first and second conditions coincide with the natural way to prove inductively that an automaton simulates another by starting with the initial state. The third condition ensures that the refinement relation prevents the introduction of deadlocks.

@@

\begin{figure}[!tb]
 \centering
   \includegraphics[width=.9\textwidth]{Figures/databaseRefine.pdf}
   \caption{The refined Open Automaton of the automaton of Fig.\ref{Fig:SpecOA}.  Two other actions,  in addition to  \texttt{qry} and \texttt{upd},  \texttt{back} representing a request to make a backup copy and    \texttt{cpy} is used to reflect the actual operation of copying. \label{Fig:RefineOA}}
\end{figure}

The refinement relation  also specifies that the refined process must follow the behaviour of the specification, not only at every compatible state but also along any scenario of the specification.
We will show in Section~\ref{sec:prop} that this refinement relation has good properties in terms of transitivity, compositionality, and reflexivity.

@@

@@

\TODO{[LUDO] relire et virer le rouge}
This definition leads to a notion of simulation as refinement the same as the one used in LOTOS \cite{Brinksma:87} and presented under the name extension refinement.
The extension consists in adding new functionalities by preserving existing ones. Indeed, in case there is no hole and no guard, i.e., in case the definitions of open automata coincide with those of automata, the relation allows new behaviour to be added in terms of new traces, but doesn't allow new deadlocks to
be added, i.e., for traces of the specification (abstract automaton) the implementation (refined automaton) must not add
deadlocks.

@@

\begin{lemma}[Tracked holes] By construction, if an automaton is the refinement of another one, it is also a refinement by tracking less holes.
\[\wrel{A_1}{A_2}{H} \land H'\subseteq H \implies \wrel{A_1}{A_2}{H'}\]
\end{lemma}

@@

\TODO{[RABEA] figure}
\begin{center}
\includegraphics[scale=0.7]{Figures/Theorem2bis}
\end{center}

@@

\TODO{[LUDO] proof sketch}

\TODO{[RABEA] example : composition and refinement}

% TODO: why is there several ovoid stacked only on one side?
%\begin{center}
%\includegraphics[scale=0.7]{Figures/Theorem1}
%\end{center}

@@

\TODO{[RABEA] RW}

@@

Additionally, because  the domains of the substitution function are disjoint, the substitution function has an effect only on the related elements,  we get:
\begin{align*}
& \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} 
%\wedge \mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}
\\
\implies & R\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}}  \land \mpar{\,\,\bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }\\
\implies & \mpar{R\mpar{s_{01}, s_{02}} \wedge  \bigwedge_\subbox{v_3\in V_3}  v^1_3=v^2_3\wedge s^1_3 = s^2_3} \psubst{ \sigma_{01} \uplus \sigma_{02} \uplus \sigma^1_{03}\uplus \sigma^2_{03}\uplus \sigma_{013} }\\
\implies & \sigma_{013} \uplus \sigma_{023} \vdash R'\mpar{s_{013}, s_{023}}
\end{align*}
\item Second, we need to prove for any open transition $t_{13}$ in $T_{13}$ originating from $s_{13}$:
\begin{align*}
		&  \OTx{13}{}{13}{13} \in \fOT{s_{13}}
\end{align*}		
there exists an indexed family $t_{23x}$ of OTs originating from $s_{23}$ that simulate it: 
\begin{align*}
		&  \mpar{\OTx{23}{x}{23x}{23x} \in \fOT{s_{23}}}^{x \in X}
\end{align*}	
\begin{align*}		
		& \text{such that } \mpar{\forall x \in X, J'_{23x} \cap H' = J'_{13} \cap H'} \text{ and }\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash R'\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H'} \beta_{13j} = \beta_{23xj} \wedge\\[12pt]
			 g_{23x} \wedge R'\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}
Recall that by definition of composition and open automata refinement we have: 
\begin{align*}
&V_{13}= V_1\uplus V^1_3 \text{ and }
V_{23}= V_2\uplus V^2_3\\
& H' \subseteq J_3\uplus \mpar{J_1 \cap J_2} \setminus \{k\} = \mpar{J_3\uplus J_1 \setminus \{k\} } \cap \mpar{J_3\uplus J_2 \setminus \{k\}} \\
\end{align*}
First of all, we have by hypothesis $\wrel{A_1}{A_2}{H}$, then for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{2}$: 
\begin{align*}
\mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_{2}}}^{x \in X} 
\end{align*}
such that $\forall x \in X, J'_{2x} \cap H = J'_1 \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\ \mpar{R\mpar{s_1, s_{2}} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}}  \qquad \quad (*)
\end{multline*}	
\smallskip

@@

We reason on the way the open transition $t_{23}$ has been obtained: either the hole $k$ of $A_2$ is involved in the transition or not.

@@

\item Second, we need to prove for any open transition $t_{12}$ in $T_{12}$ originating from $s_{12}$

@@

We reason on the way the open transition $t_{13}$ has been obtained: either the hole $k$ of $A_1$ is involved in the transition or not.
